---
title: "Architecture Overview"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, dplyr::rename, .quiet = TRUE)
```

## Class hierarchy

PKNCA uses four S3 classes: `PKNCAconc` and `PKNCAdose` feed `PKNCAdata`, and `pk.nca()` wraps the result into `PKNCAresults`:

```{mermaid}
classDiagram
    class PKNCAconc {
        data: data.frame
        formula: conc~time|groups
        subject: column name
        sparse: logical
        units: concu/timeu
        exclude / exclude_half.life
    }
    class PKNCAdose {
        data: data.frame
        formula: dose~time|groups
        route: intravas./extravas.
        duration / rate
        units: doseu/timeu
    }
    class PKNCAdata {
        conc: PKNCAconc
        dose: PKNCAdose
        intervals: data.frame
        impute: method string
        units: pknca_units_table()
        options: named list
    }
    class PKNCAresults {
        result: data.frame
        data: PKNCAdata
    }

    PKNCAconc --> PKNCAdata
    PKNCAdose --> PKNCAdata
    PKNCAdata --> PKNCAresults
```

## The interval column system

The parameter registry is the core extensibility mechanism. Every NCA parameter is registered as an **interval column** via `add.interval.col()`.

```{r}
# Inspect registration for auclast
cols <- get.interval.cols()
str(cols[["auclast"]])
```

Each entry contains:

| Field | Purpose |
|---|---|
| `FUN` | Name of the function that computes this parameter (character string, or `NA` for parameters populated by another function's multi-column result) |
| `values` | Allowed values in the interval column (typically `c(FALSE, TRUE)`) |
| `unit_type` | Unit category used for unit assignment and conversion |
| `pretty_name` | Human-readable name used in summary tables |
| `depends` | Character vector of parameters that must be computed first |
| `desc` | Human-readable description |
| `sparse` | Whether this parameter works for sparse PK |
| `formalsmap` | Maps function arguments to interval/group/options data |
| `datatype` | Calculation type; only `"interval"` is currently supported |
| `pptestcd_cdisc` | CDISC PPTESTCD code (≥ 0.12.2) |
| `pptest_cdisc` | CDISC PPTEST label (≥ 0.12.2) |

## formalsmap

`formalsmap` is how PKNCA passes the right data to each parameter function. Arguments can be sourced from four places:

| Source type | Example | What it provides |
|---|---|---|
| Interval column | `conc`, `time` | Concentrations/times for the current interval |
| Other parameters | `auclast = "auclast"` | A previously computed NCA result |
| Interval bounds | `start`, `end` | The interval start/end times |
| Options | `options` | The full analysis options list |

Individual option names (e.g. `auc.method`) cannot be mapped through `formalsmap`; only the whole `options` list can, and parameter functions read specific settings from it via `PKNCA.choose.option()`.

## Writing a custom parameter

Registering a new parameter is a single `add.interval.col()` call that names the computing function; PKNCA then derives execution order from `depends`, assigns units from `unit_type`, and collects results automatically. A minimal toy registration:

```{r}
# Toy illustration: ratio of two built-in parameters
pk.calc.demo.ratio <- function(cmax, cmin) {
  cmax / cmin
}

add.interval.col(
  name        = "demo.ratio",
  FUN         = "pk.calc.demo.ratio",  # must be a character string, not the function object
  unit_type   = "unitless",
  pretty_name = "Demo ratio (Cmax/Cmin)",
  desc        = "Toy example: Cmax divided by Cmin",
  depends     = c("cmax", "cmin")
)

# The registry now contains it alongside the built-in parameters
str(get.interval.cols()[["demo.ratio"]])
```

For the full walkthrough — parameters computed from raw concentration-time data, multi-output functions, and custom summary statistics — see the [Writing Custom Parameters chapter](../user/custom-parameters.qmd) and the package vignette [Writing PKNCA Parameter Functions](https://humanpred.github.io/pknca/articles/v80-writing-parameter-functions.html).

## Options system

PKNCA's options are stored in a package-level environment; `PKNCA.options()` returns them as a named list. It is accessed in several ways:

```{r}
# 1. Read all options
PKNCA.options()

# 2. Read a specific option
PKNCA.options("min.hl.points")

# 3. Set globally — save the old value first so you can restore it
old_val <- PKNCA.options("min.hl.points")[[1]]
PKNCA.options(min.hl.points = 4)
PKNCA.options("min.hl.points")

# 4. Restore by passing the saved value as a named argument
PKNCA.options(min.hl.points = old_val)
PKNCA.options("min.hl.points")
```

> Prefer passing `options = list(...)` to `PKNCAdata()` over mutating global options. Global changes persist across analyses in the same session.

## Sparse PK

Sparse PK (one or two samples per subject, common in preclinical or pediatric studies) is handled via a parallel code path:

- Set `sparse = TRUE` in `PKNCAconc()`
- Concentrations are pooled across subjects per timepoint
- Mean concentration-time profile is computed, then NCA is run on that profile
- Supported parameters are flagged with `sparse = TRUE` in their `add.interval.col()` registration

For the full workflow, see the [Sparse NCA chapter](../user/sparse.qmd) and the vignette [Sparse NCA Calculations](https://humanpred.github.io/pknca/articles/v04-sparse.html).

## Key functions

All of these are exported except where marked internal:

| Function | Purpose |
|---|---|
| `get.interval.cols()` | Return full parameter registry |
| `PKNCA:::sort_interval_cols()` | Topological sort → execution order (internal, unexported) |
| `pk.nca.interval()` | Compute all parameters for one interval/group |
| `pknca_units_table()` | Build unit assignment/conversion table |
| `PKNCA_impute_method_*()` | Built-in imputation methods |
| `pk.calc.auc.last()` | Core AUC calculation (trapezoidal) |
| `pk.calc.half.life()` | λz regression with best-fit selection |

---

> This overview covers the surface. The business rules (`002-pk.business.rules.R`) and the validation framework ([`v60-PKNCA-validation.Rmd`](https://humanpred.github.io/pknca/articles/v60-PKNCA-validation.html)) in the PKNCA sources are richer topics worth exploring further.


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html) · [add.interval.col()](https://humanpred.github.io/pknca/reference/add.interval.col.html) · [get.interval.cols()](https://humanpred.github.io/pknca/reference/get.interval.cols.html) · [PKNCA.options()](https://humanpred.github.io/pknca/reference/PKNCA.options.html) · [pknca_units_table()](https://humanpred.github.io/pknca/reference/pknca_units_table.html) · [pk.nca.interval()](https://humanpred.github.io/pknca/reference/pk.nca.interval.html) · [PKNCA.choose.option()](https://humanpred.github.io/pknca/reference/PKNCA.choose.option.html) · [pk.calc.half.life()](https://humanpred.github.io/pknca/reference/pk.calc.half.life.html)
:::
