---
title: "NCA Parameter Dependencies"
---

```{r setup, include=FALSE}
library(PKNCA)
library(visNetwork)
library(dplyr)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

PKNCA resolves parameter dependencies automatically: when you request `aucinf.obs`, PKNCA internally ensures `lambda.z` is computed first because `aucinf.obs` depends on it. These dependencies are declared at registration time — every NCA parameter is registered via `add.interval.col()` with a `depends` vector naming the parameters that must be computed before it, and the full registry is accessible via `get.interval.cols()`. The graph below makes the whole dependency network explicit.

## Interactive dependency graph

```{r build-graph}
# Get all registered interval columns (parameters)
cols <- get.interval.cols()

# Build edge list from the 'depends' field
edges <- lapply(names(cols), function(param) {
  deps <- cols[[param]]$depends
  if (length(deps) == 0) return(NULL)
  data.frame(from = deps, to = param, stringsAsFactors = FALSE)
}) |>
  bind_rows()

# All unique nodes
all_nodes <- unique(c(edges$from, edges$to, names(cols)))

# Assign category for coloring
categorize <- function(name) {
  case_when(
    grepl("^auc",       name) ~ "AUC",
    grepl("^lambda|half\\.life|r\\.squared", name) ~ "Half-life / λz",
    grepl("^cl\\.|^vz\\.|^vss", name) ~ "CL / Volume",
    grepl("^mrt",       name) ~ "MRT",
    grepl("^c(max|min|0|last|first)|^t(max|min|last|first|lag)|clast", name) ~ "Cmax / Tmax",
    grepl("^f$|bioavail|^ae|^fe|^clr|^vd",  name) ~ "Bioavailability / Renal",
    grepl("^tss",       name) ~ "Steady-state",
    TRUE ~ "Other"
  )
}

category_colors <- c(
  "AUC"                    = "#4e79a7",
  "Half-life / λz"         = "#f28e2b",
  "CL / Volume"            = "#e15759",
  "MRT"                    = "#76b7b2",
  "Cmax / Tmax"            = "#59a14f",
  "Bioavailability / Renal"= "#edc948",
  "Steady-state"           = "#b07aa1",
  "Other"                  = "#aaaaaa"
)

nodes <- data.frame(
  id    = all_nodes,
  label = all_nodes,
  group = categorize(all_nodes),
  stringsAsFactors = FALSE
)

graph <- visNetwork(nodes, edges,
           width = "100%", height = "700px",
           main = "PKNCA NCA Parameter Dependencies") |>
  visNodes(shape = "dot", size = 14,
           font = list(size = 12),
           borderWidth = 2) |>
  visEdges(arrows = "to",
           color  = list(color = "#aaaaaa", highlight = "#333333"),
           smooth = list(type = "cubicBezier")) |>
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 2, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, main = "Select parameter")
  ) |>
  visLegend(useGroups = TRUE, position = "right") |>
  visPhysics(stabilization = list(iterations = 200)) |>
  visLayout(randomSeed = 42)

# Register each category as a visNetwork group: nodes and legend swatches both
# take their color from the group definition, so they always match
for (grp in names(category_colors)) {
  graph <- visGroups(graph, groupname = grp, color = category_colors[[grp]])
}
graph
```

**How to use this graph:**

- Click a parameter node to highlight its immediate dependencies and dependents
- The "Select parameter" dropdown jumps to a specific node
- Drag nodes to rearrange; scroll to zoom
- Colors indicate parameter family (see legend)

## Dependency table

For a text-based reference:

```{r dep-table}
dep_table <- lapply(names(cols), function(param) {
  data.frame(
    parameter = param,
    depends_on = if (length(cols[[param]]$depends) == 0) "(none)"
                 else paste(cols[[param]]$depends, collapse = ", "),
    stringsAsFactors = FALSE
  )
}) |>
  bind_rows() |>
  arrange(parameter)

knitr::kable(dep_table, col.names = c("Parameter", "Depends on"))
```

## Parameter-to-function mapping

Every interval parameter is computed by a function registered alongside it — usually named `pk.calc.*()` — and the registry entry returned by `get.interval.cols()` records that function name in its `$FUN` field, together with the parameter's unit type and description. Where `$FUN` is `NA`, the parameter has no dedicated function: it is either produced as part of another function's output (for example, `r.squared` and `lambda.z.time.first` are returned by the same half-life fit that produces `half.life`) or it is an interval boundary (`start`, `end`).

The table below is the complete index of the calculation API: every registered parameter, the function that computes it, its unit type, and its description.

```{r fun-map}
cols <- get.interval.cols()

fun_map <- data.frame(
  parameter = names(cols),
  # $FUN is a character function name, or logical NA when the parameter has no
  # dedicated function; as.character() coerces both to character
  fun       = vapply(cols, function(x) as.character(x$FUN), character(1)),
  unit_type = vapply(cols, function(x) x$unit_type, character(1)),
  desc      = vapply(cols, function(x) x$desc, character(1)),
  stringsAsFactors = FALSE
) |>
  arrange(parameter)

knitr::kable(
  fun_map,
  row.names = FALSE,
  col.names = c("Parameter", "Function", "Unit type", "Description")
)
```

## Dependency resolution

When you call `pk.nca()`, PKNCA sorts the registry internally with `PKNCA:::sort_interval_cols()` (unexported) to produce a topologically sorted execution order — parameters with no dependencies run first, then those that depend on already-computed values.

```{r sort-order}
# Count the direct dependencies of each parameter
cols <- get.interval.cols()
dep_counts <- sapply(cols, function(x) length(x$depends))
all_deps <- data.frame(
  parameter  = names(dep_counts),
  n_depends  = unname(dep_counts)
) |>
  arrange(n_depends, parameter)

cat(sprintf("%d parameters have no dependencies (base parameters).\n", sum(all_deps$n_depends == 0)))
cat(sprintf("%d parameters have at least one dependency (derived parameters).\n\n", sum(all_deps$n_depends > 0)))

# Show only derived parameters
all_deps |>
  filter(n_depends > 0) |>
  knitr::kable(
    col.names = c("Parameter", "N dependencies"),
    caption   = "Derived parameters ordered by number of direct dependencies"
  )
```

## Querying dependents: `get.parameter.deps()`

`get.parameter.deps()` is the reverse lookup: given a parameter name, it returns that parameter and every parameter downstream of it — the ones whose results would be affected if you excluded or modified it.

```{r parameter-deps}
get.parameter.deps("auclast")
```


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [get.interval.cols()](https://humanpred.github.io/pknca/reference/get.interval.cols.html) · [add.interval.col()](https://humanpred.github.io/pknca/reference/add.interval.col.html) · [get.parameter.deps()](https://humanpred.github.io/pknca/reference/get.parameter.deps.html)
:::
