---
title: "Writing Custom Parameter Functions"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)

# Shared dataset
d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)

o_conc <- PKNCAconc(d_conc, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")
```

## The parameter registry

Every NCA parameter that PKNCA can compute is **registered** in an internal registry via `add.interval.col()`. When you request `auclast = TRUE` in an interval, PKNCA looks up the registered function for `auclast` and calls it with the right arguments.

You can extend the registry with your own parameters — new functions that compute anything derivable from concentrations, times, doses, or other already-computed parameters. See also the package vignette [Writing PKNCA Parameter Functions](https://humanpred.github.io/pknca/articles/v80-writing-parameter-functions.html).

```{r}
# Number of parameters registered by default
length(get.interval.cols())
```

---

## Anatomy of `add.interval.col()`

```r
add.interval.col(
  name        = "my.param",         # column name used in intervals data frame
  FUN         = "pk.calc.my.param", # character string: function name
  values      = c(FALSE, TRUE),     # allowed values in intervals (FALSE = skip, TRUE = compute)
  unit_type   = "time",             # units category (see below)
  pretty_name = "My parameter",     # human-readable label
  desc        = "What it does",     # one-line description
  depends     = NULL,               # other parameters that must be computed first
  sparse      = FALSE,              # TRUE if this is a sparse-PK parameter
  formalsmap  = list(),             # list(<your function arg> = "<PKNCA data name>")
  datatype    = "interval",         # "interval" (default), "individual", or "population"
                                    #   (only "interval" is currently supported; the others error)
  pptestcd_cdisc = "MYPARAM",       # CDISC PPTESTCD code for the parameter (≥ 0.12.2)
  pptest_cdisc   = "My Parameter"   # CDISC PPTEST label for the parameter (≥ 0.12.2)
)
```

**Key rule:** `FUN` must be a **character string** (the function name), not a function object.

### Available `unit_type` values

`"unitless"`, `"fraction"`, `"%"`, `"count"`, `"time"`, `"inverse_time"`, `"amount"`, `"amount_dose"`, `"amount_time"`, `"conc"`, `"conc_dosenorm"`, `"dose"`, `"volume"`, `"auc"`, `"aumc"`, `"auc_dosenorm"`, `"aumc_dosenorm"`, `"clearance"`, `"renal_clearance"`, `"renal_clearance_dosenorm"`

---

## The `formalsmap` argument

Each `formalsmap` entry has the form `<your function's argument> = "<PKNCA data name>"` — the **names** of the list must be your function's formal arguments, and the **values** are the PKNCA data or parameter names that supply them. For example, PKNCA registers `mrt.obs` with `formalsmap = list(auc = "aucinf.obs", aumc = "aumcinf.obs")`: the computed `aucinf.obs` value is passed to the function's `auc` argument.

Standard PKNCA data names available as `formalsmap` values:

| PKNCA name | What it provides |
|---|---|
| `conc` | Concentration vector for the current interval |
| `time` | Time vector for the current interval |
| `dose` | Total dose in the current interval |
| `start` | Interval start time |
| `end` | Interval end time |
| `half.life` | Computed half-life (if in `depends`) |
| `auclast` | Computed AUClast (if in `depends`) |
| `cmax` | Computed Cmax (if in `depends`) |
| *(any parameter name)* | Any other registered parameter in `depends` |

If your function argument names **exactly match** PKNCA's standard names, `formalsmap` can be omitted or left empty.

---

## Example 1: derived from other parameters

A parameter that divides two already-computed parameters. Use `depends` to declare what must be computed first (shown here without `formalsmap` because the function's argument names match the parameter names).

```{r}
# Exposure ratio: AUClast / Cmax  (units: time, since AUC/conc = time × conc/conc = time)
pk.calc.exposure.ratio <- function(auclast, cmax) {
  auclast / cmax
}

add.interval.col(
  name        = "exposure.ratio",
  FUN         = "pk.calc.exposure.ratio",
  values      = c(FALSE, TRUE),
  unit_type   = "time",
  pretty_name = "Exposure ratio (AUClast / Cmax)",
  desc        = "AUClast divided by Cmax",
  depends     = c("auclast", "cmax")
)
```

```{r}
o_nca_er <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start          = 0, end = Inf,
  auclast        = TRUE,
  cmax           = TRUE,
  exposure.ratio = TRUE
)))

as.data.frame(o_nca_er) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "exposure.ratio")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(check = auclast / cmax) |>
  arrange(subject)
```

---

## Example 2: computed from raw concentration-time data

A parameter that accesses `conc` and `time` directly. PKNCA passes the raw vectors from the concentration data. No `depends` needed. The identity `formalsmap` below is shown for explicitness — because the argument names match PKNCA's standard names, it could equally be omitted.

```{r}
# Concentration at exactly 2 hours (interpolated)
pk.calc.c_at_2h <- function(conc, time) {
  interp.extrap.conc(conc = conc, time = time, time.out = 2)
}

add.interval.col(
  name        = "c.2h",
  FUN         = "pk.calc.c_at_2h",
  values      = c(FALSE, TRUE),
  unit_type   = "conc",
  pretty_name = "Concentration at 2 h",
  desc        = "Interpolated concentration at t = 2 h",
  depends     = NULL,
  formalsmap  = list(conc = "conc", time = "time")
)
```

```{r}
o_nca_c2 <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start = 0, end = Inf, c.2h = TRUE
)))

as.data.frame(o_nca_c2) |>
  filter(PPTESTCD == "c.2h") |>
  select(subject, PPORRES) |>
  arrange(subject)
```

---

## Example 3: multiple related parameters sharing a dependency

When computing several related quantities (e.g. early AUC, late AUC, and their ratio), register each as its own function. The late-AUC and fraction functions read the already-computed `auclast` dependency.

```{r}
pk.calc.auc.early <- function(conc, time) {
  t_idx <- time <= 4
  if (sum(t_idx) < 2) return(NA_real_)
  pk.calc.auc.last(conc[t_idx], time[t_idx])
}

pk.calc.auc.late <- function(conc, time, auclast) {
  t_idx <- time <= 4
  if (sum(t_idx) < 2) return(NA_real_)
  auclast - pk.calc.auc.last(conc[t_idx], time[t_idx])
}

pk.calc.auc.early.frac <- function(conc, time, auclast) {
  t_idx <- time <= 4
  if (sum(t_idx) < 2) return(NA_real_)
  pk.calc.auc.last(conc[t_idx], time[t_idx]) / auclast
}

add.interval.col(
  name        = "auc.early",
  FUN         = "pk.calc.auc.early",
  values      = c(FALSE, TRUE),
  unit_type   = "auc",
  pretty_name = "AUC to last sample <= 4h",
  desc        = "AUC to the last sample at or before 4 h",
  depends     = NULL,
  formalsmap  = list(conc = "conc", time = "time")
)

add.interval.col(
  name        = "auc.late",
  FUN         = "pk.calc.auc.late",
  values      = c(FALSE, TRUE),
  unit_type   = "auc",
  pretty_name = "AUC 4h to last",
  desc        = "AUC from 4 h to last sample",
  depends     = "auclast",
  formalsmap  = list(conc = "conc", time = "time", auclast = "auclast")
)

add.interval.col(
  name        = "auc.early.fraction",
  FUN         = "pk.calc.auc.early.frac",
  values      = c(FALSE, TRUE),
  unit_type   = "fraction",
  pretty_name = "Early AUC fraction",
  desc        = "Fraction of AUClast occurring in first 4 h",
  depends     = "auclast",
  formalsmap  = list(conc = "conc", time = "time", auclast = "auclast")
)
```

> **Note:** One function *can* populate several output parameters at once by returning a **one-row data frame** whose column names are registered parameter names — `pk.nca()` saves every column of the returned data frame. This is exactly how `pk.calc.half.life()` populates `half.life`, `lambda.z`, `r.squared`, and the other half-life diagnostics. Each output still needs its own `add.interval.col()` registration: register the function on the primary parameter, and register the other columns with `FUN = NA` plus a `depends` on the primary (`formalsmap` may not be given when `FUN = NA`). The one-function-per-parameter pattern shown above is the simpler option when each output is cheap to compute.

```{r}
o_nca_split <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start              = 0, end = Inf,
  auclast            = TRUE,
  auc.early          = TRUE,
  auc.late           = TRUE,
  auc.early.fraction = TRUE
)))

as.data.frame(o_nca_split) |>
  filter(PPTESTCD %in% c("auclast", "auc.early", "auc.late", "auc.early.fraction")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

The same split can be computed by **one** multi-output function returning a one-row data frame:

```{r}
pk.calc.auc.split <- function(conc, time, auclast) {
  t_idx <- time <= 4
  early <- if (sum(t_idx) < 2) NA_real_ else pk.calc.auc.last(conc[t_idx], time[t_idx])
  data.frame(auc.early.df = early, auc.late.df = auclast - early)
}

add.interval.col(
  name        = "auc.early.df",
  FUN         = "pk.calc.auc.split",  # computes BOTH auc.early.df and auc.late.df
  values      = c(FALSE, TRUE),
  unit_type   = "auc",
  pretty_name = "AUC 0-4h (multi-output)",
  desc        = "AUC from 0 to 4 h, from a multi-output function",
  depends     = "auclast",
  formalsmap  = list(conc = "conc", time = "time", auclast = "auclast")
)

add.interval.col(
  name        = "auc.late.df",
  FUN         = NA,                   # populated by pk.calc.auc.split via auc.early.df
  values      = c(FALSE, TRUE),
  unit_type   = "auc",
  pretty_name = "AUC 4h to last (multi-output)",
  desc        = "AUC from 4 h to last sample, from a multi-output function",
  depends     = "auc.early.df"
)

o_nca_multi <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start = 0, end = Inf,
  auclast = TRUE, auc.early.df = TRUE, auc.late.df = TRUE
)))

as.data.frame(o_nca_multi) |>
  filter(PPTESTCD %in% c("auclast", "auc.early.df", "auc.late.df")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

---

## Inspecting registered parameters

```{r}
# Confirm your parameter is registered
cols <- get.interval.cols()
cat("exposure.ratio registered:", "exposure.ratio" %in% names(cols), "\n")
cat("c.2h registered:          ", "c.2h" %in% names(cols), "\n")
```

```{r}
# Inspect registration details
str(cols[["exposure.ratio"]])
```

---

## Custom summary statistics for your parameter

After registering the parameter, set a custom summary rule with `PKNCA.set.summary()`:

```{r}
PKNCA.set.summary(
  "exposure.ratio",
  description = "geometric mean [CV%]",
  point  = PKNCA::geomean,
  spread = PKNCA::geocv
)

summary(pk.nca(PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start = 0, end = Inf,
  auclast = TRUE, cmax = TRUE, exposure.ratio = TRUE
))))
```

The vignette's [summary section](https://humanpred.github.io/pknca/articles/v80-writing-parameter-functions.html#tell-pknca-how-to-summarize-the-parameter) covers the same registration.

---

## Important notes

- **Registration is session-scoped.** Custom parameters registered with `add.interval.col()` persist for the life of your R session. Re-run registration code in each session (e.g. in your analysis script or package).
- **`FUN` must be a character string.** Passing a function object causes an error.
- **`depends` controls evaluation order.** List all parameters your function reads. PKNCA resolves the dependency graph and ensures they are computed before your function is called.
- **Functions must be in scope** when `pk.nca()` runs. Define them in the global environment or in a package loaded before calling `pk.nca()`.


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html) · [add.interval.col()](https://humanpred.github.io/pknca/reference/add.interval.col.html) · [get.interval.cols()](https://humanpred.github.io/pknca/reference/get.interval.cols.html) · [interp.extrap.conc()](https://humanpred.github.io/pknca/reference/interp.extrap.conc.html) · [pk.calc.auc.last()](https://humanpred.github.io/pknca/reference/pk.calc.auxc.html) · [PKNCA.set.summary()](https://humanpred.github.io/pknca/reference/PKNCA.set.summary.html) · [geomean()](https://humanpred.github.io/pknca/reference/geomean.html) · [geocv()](https://humanpred.github.io/pknca/reference/geomean.html)
:::
