---
title: "Concentration Imputation"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## Why imputation is needed

NCA intervals are defined by a `start` and `end` time. For AUC and related
parameters, PKNCA needs a concentration at the exact interval start. Two common
situations require imputation:

1. **No t = 0 sample.** The earliest measured concentration is after dosing, so
   there is no observed value at the interval start (time 0).
2. **Pre-dose sample.** A concentration was measured before the nominal dose
   time (e.g. a residual carry-over sample). The interval starts at time 0 but
   the only nearby observation has a negative relative time.

Imputation is applied just before each individual interval is calculated, on a
per-subject, per-interval basis. The original data frame is never modified.

### Example data: Theoph with t = 0 removed

```{r}
Theoph_df <- as.data.frame(Theoph)

# Remove the t = 0 observation to create a missing-baseline scenario
d_conc_no0 <- Theoph_df |> filter(Time > 0)
d_dose      <- Theoph_df |>
  filter(Time == 0) |>
  mutate(amt = Dose * Wt)

# Without imputation there is no concentration at the interval start
o_conc <- PKNCAconc(d_conc_no0, conc ~ Time | Subject)
o_dose <- PKNCAdose(d_dose, amt ~ Time | Subject)
```

```{r}
# What does subject 1's profile look like?
d_conc_no0 |>
  filter(Subject == "1") |>
  head(5)
```

---

## The three built-in strategies

Verified with `args()`:

```r
args(PKNCA_impute_method_start_conc0)
# function(conc, time, start = 0, ..., options = list())

args(PKNCA_impute_method_start_predose)
# function(conc, time, start, end, conc.group, time.group, ...,
#          max_shift = NA_real_, options = list())

args(PKNCA_impute_method_start_cmin)
# function(conc, time, start, end, ..., options = list())
```

| Name | String to use | What it does |
|------|--------------|--------------|
| `PKNCA_impute_method_start_conc0` | `"start_conc0"` | Inserts a concentration of 0 at the interval start time |
| `PKNCA_impute_method_start_predose` | `"start_predose"` | Uses the last observed concentration before the interval start |
| `PKNCA_impute_method_start_cmin` | `"start_cmin"` | Uses the minimum concentration observed within the interval |

### `start_conc0` — zero concentration at interval start

Appropriate in a single-dose setting when no drug is expected before dosing.
The drug concentration at time 0 is assumed to be 0.

```{r}
o_data_c0 <- PKNCAdata(o_conc, o_dose, impute = "start_conc0")
o_data_c0$impute   # verify the impute string is stored
```

```{r}
res_c0 <- pk.nca(o_data_c0)

as.data.frame(res_c0) |>
  filter(Subject == "1",
                PPTESTCD %in% c("auclast", "cmax", "tmax")) |>
  select(Subject, PPTESTCD, PPORRES)
```

### `start_predose` — last pre-dose concentration

Appropriate in multiple-dose or steady-state settings where a true pre-dose
(trough) sample was collected just before administration. The method looks
outside the current interval for the most recent observed concentration. The
`max_shift` argument caps how far forward that pre-dose sample may be shifted
to the interval start: it defaults to 5% of the interval duration
(`0.05 * (end - start)`), or, when `end` is infinite, 5% of the time from
`start` to the last sample.

```{r}
o_data_pd <- PKNCAdata(o_conc, o_dose, impute = "start_predose")
o_data_pd$impute
```

### `start_cmin` — minimum concentration in interval

Appropriate as a conservative fallback: if no pre-dose value is available and
zero cannot be assumed (e.g. metabolite or endogenous compound), use the
minimum observed concentration in the interval as a lower-bound estimate.

```{r}
o_data_cmin <- PKNCAdata(o_conc, o_dose, impute = "start_cmin")
o_data_cmin$impute
```

### The three strategies side by side

Running all three on subject 1 (whose first sample is at t = 0.25) makes the
differences concrete:

```{r}
d1_conc <- d_conc_no0 |> filter(Subject == "1")
d1_dose <- d_dose |> filter(Subject == "1")
o_conc1 <- PKNCAconc(d1_conc, conc ~ Time | Subject)
o_dose1 <- PKNCAdose(d1_dose, amt ~ Time | Subject)
ivl1    <- data.frame(start = 0, end = Inf, auclast = TRUE)

auclast_with <- function(method) {
  res <- as.data.frame(pk.nca(PKNCAdata(o_conc1, o_dose1, intervals = ivl1,
                                        impute = method)))
  data.frame(impute = method, auclast = res$PPORRES[res$PPTESTCD == "auclast"])
}

bind_rows(auclast_with("start_conc0"),
                 auclast_with("start_predose"),
                 auclast_with("start_cmin"))
```

`start_conc0` adds conc = 0 at t = 0, while `start_cmin` adds the minimum
concentration observed in the interval (here 2.84 mg/L, the first sample) —
so `start_cmin` gives the slightly larger AUC. `start_predose` looks for a
sample at or before the interval start, and this dataset has none: it imputes
nothing, the interval still has no concentration at its start time, and
`auclast` is `NA` (the warning printed by the chunk explains why).

---

## Applying imputation globally

The single-strategy examples above already used the global form: pass a method
name string to the `impute` argument of `PKNCAdata()`, and the same strategy is
applied to every subject and every interval — the side-by-side comparison
restricted itself to subject 1 only to keep the output small.

```{r}
o_data_global <- PKNCAdata(o_conc, o_dose, impute = "start_conc0")

res_global <- pk.nca(o_data_global)
df_global  <- as.data.frame(res_global)

# auclast is populated for all subjects despite the missing t = 0
df_global |>
  filter(PPTESTCD == "auclast") |>
  select(Subject, PPORRES) |>
  head(6)
```

---

## Chaining methods

The side-by-side comparison left `start_predose` with an `NA` because this
dataset has no pre-dose sample. Chaining resolves exactly that situation:
provide a comma-separated string to apply methods in sequence. Methods are
applied in order; a later method has no effect if an earlier one already
added a point at the interval start.

```{r}
# Try start_predose first; if it cannot find a pre-dose observation,
# fall back to inserting zero
o_data_chain <- PKNCAdata(o_conc, o_dose,
                          impute = "start_predose,start_conc0")
o_data_chain$impute
```

```{r}
res_chain <- pk.nca(o_data_chain)

as.data.frame(res_chain) |>
  filter(Subject == "1", PPTESTCD == "auclast") |>
  select(Subject, PPTESTCD, PPORRES)
```

For subject 1 the chained result matches the global `start_conc0` run in the
previous section: `start_predose` found no pre-dose sample to carry forward,
so the fallback inserted the zero.

---

## Per-interval imputation

Different intervals in the same dataset can use different strategies. Add an
`impute` column to the intervals data frame, then pass the column name (as a
string) to the `impute` argument of `PKNCAdata()`.

```{r}
my_intervals <- data.frame(
  start  = c(0,   0),
  end    = c(24,  Inf),
  auclast = c(TRUE,  FALSE),
  aucinf.obs = c(FALSE, TRUE),
  cmax   = c(TRUE,  TRUE),
  impute = c("start_conc0", "start_predose,start_conc0")
)

o_data_per_interval <- PKNCAdata(
  o_conc, o_dose,
  intervals = my_intervals,
  impute    = "impute"         # name of the column to read
)

o_data_per_interval$impute
o_data_per_interval$intervals[, c("start", "end", "impute")]
```

```{r}
res_per <- pk.nca(o_data_per_interval)

as.data.frame(res_per) |>
  filter(Subject == "1",
                PPTESTCD %in% c("auclast", "aucinf.obs", "cmax")) |>
  select(Subject, PPTESTCD, PPORRES)
```

---

## Writing a custom imputation function

### Naming convention

The function name **must** follow the pattern `PKNCA_impute_method_<name>`.
PKNCA prepends `PKNCA_impute_method_` to the string you supply and looks up
that single function on the search path, so the convention is mandatory.

### Required signature

Modeled after the built-in functions:

```r
# Minimal required arguments (from args(PKNCA_impute_method_start_conc0)):
PKNCA_impute_method_<name> <- function(conc, time, start = 0, ..., options = list())
```

The function receives:

| Argument | Description |
|----------|-------------|
| `conc` | Numeric vector of concentrations for the current subject/interval |
| `time` | Numeric vector of times (same length as `conc`) |
| `start` | Interval start time |
| `...` | Additional arguments (must be accepted and ignored) |
| `options` | Named list of PKNCA options |

### Return value

A data frame with exactly two columns, `conc` and `time`, sorted by `time`.
It should include all original observations **plus** any imputed points.

### Example: impute with the mean of the first two observations

```{r}
PKNCA_impute_method_start_mean2 <- function(conc, time, start = 0, ...,
                                             options = list()) {
  # Only impute if the interval start is not already observed
  if (!start %in% time) {
    # Use the mean of the first two observed concentrations
    imputed_conc <- mean(conc[order(time)][seq_len(min(2, length(conc)))])
    conc <- c(imputed_conc, conc)
    time <- c(start, time)
  }
  result <- data.frame(conc = conc, time = time)
  result[order(result$time), ]
}
```

```{r}
# Use the custom method by its suffix
o_data_custom <- PKNCAdata(o_conc, o_dose, impute = "start_mean2")
res_custom     <- pk.nca(o_data_custom)

as.data.frame(res_custom) |>
  filter(Subject == "1", PPTESTCD == "auclast") |>
  select(Subject, PPTESTCD, PPORRES)
```

---

## When to use each strategy

| Situation | Recommended strategy |
|-----------|----------------------|
| Single-dose study, no pre-dose drug expected | `start_conc0` |
| Multiple-dose or steady-state, trough sample collected | `start_predose` |
| Endogenous analyte or metabolite, zero implausible | `start_cmin` |
| Uncertain: try pre-dose, fall back to zero | `"start_predose,start_conc0"` |
| Non-standard logic required | Custom `PKNCA_impute_method_*` function |

---

## Summary

- Imputation is triggered when no observation exists at the interval start time.
- The three built-in methods are referenced by their suffix strings:
  `"start_conc0"`, `"start_predose"`, `"start_cmin"`.
- Global imputation: `PKNCAdata(..., impute = "start_conc0")`.
- Chained methods: `impute = "start_predose,start_conc0"`.
- Per-interval: add an `impute` column to the intervals data frame and pass
  that column name to the `impute` argument of `PKNCAdata()`.
- Custom functions: name them `PKNCA_impute_method_<name>` and return a
  two-column data frame (`conc`, `time`).


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html) · [PKNCA_impute_method_start_conc0()](https://humanpred.github.io/pknca/reference/PKNCA_impute_method_start_conc0.html) · [PKNCA_impute_method_start_predose()](https://humanpred.github.io/pknca/reference/PKNCA_impute_method_start_predose.html) · [PKNCA_impute_method_start_cmin()](https://humanpred.github.io/pknca/reference/PKNCA_impute_method_start_cmin.html)
:::
