---
title: "Superposition"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## What superposition does

**Superposition** predicts steady-state concentration-time profiles from a single-dose profile by linearly summing lagged copies of the single-dose curve. This is the NCA equivalent of multi-dose simulation, and it requires no compartmental model. Observed steady-state parameters (Cav, PTR, fluctuation) are covered in [Multiple-Dose and Steady-State](multiple-dose.qmd); estimating when steady state is reached is covered in [Time to Steady State](tss.qmd).

The core assumption is **linear, time-invariant pharmacokinetics**: the drug does not accumulate non-linearly, and clearance and volume do not change over time.

```{mermaid}
flowchart LR
    A["Single dose<br/>profile (0→∞)"] --> B["Shift by τ,<br/>shift by 2τ, ..."]
    B --> C["Sum contributions<br/>at each time t"]
    C --> D["Predicted SS<br/>profile (0→τ)"]
```

---

## Basic superposition

`Theoph` records dose in mg/kg; converting it to each subject's total dose in mg gives the value that `dose.input` expects:

```{r}
d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)

# Theoph dose is mg/kg — multiply by body weight for the total mg dose
d_dose <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)
```

```{r}
# Use Theoph subject 1 as single-dose reference
subj1 <- d_conc |> filter(subject == "1")
actual_dose <- d_dose |> filter(subject == "1") |> pull(dose)

o_conc_1 <- PKNCAconc(subj1, conc ~ time | subject)

# Predict steady-state profile with τ = 24 h, show 3 dosing intervals
ss_profile <- superposition(
  o_conc_1,
  tau         = 24,           # dosing interval (h)
  dose.input  = actual_dose,  # dose used to generate the single-dose data
  dose.amount = actual_dose,  # dose to simulate at SS (same here)
  n.tau       = 3,            # simulate 3 intervals
  check.blq   = FALSE         # Theoph has non-zero first sample
)

head(ss_profile, 12)
```

---

## Visualizing the approach to steady state

With a finite `n.tau`, `superposition()` returns the concentration profile over the dosing interval after that many doses; with `n.tau = Inf` it accumulates doses until the profile converges to steady state (within `steady.state.tol`). Overlaying the profiles after 1, 2, and 3 doses on the converged profile makes the approach to steady state visible.

```{r}
approach <- lapply(1:3, function(k) {
  superposition(o_conc_1, tau = 24, dose.input = actual_dose,
                dose.amount = actual_dose, n.tau = k, check.blq = FALSE) |>
    mutate(profile = paste("After dose", k))
}) |> bind_rows()

ss_converged <- superposition(
  o_conc_1,
  tau         = 24,
  dose.input  = actual_dose,
  dose.amount = actual_dose,
  n.tau       = Inf,
  check.blq   = FALSE
)

bind_rows(approach, ss_converged |> mutate(profile = "Steady state")) |>
  mutate(profile = factor(profile, levels = c(paste("After dose", 1:3), "Steady state"))) |>
  ggplot(aes(x = time, y = conc, colour = profile)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Superposition: approach to steady state (τ = 24 h)",
       x = "Time within interval (h)", y = "Predicted concentration (mg/L)",
       colour = NULL) +
  theme_minimal()
```

---

## Changing dose at steady state

To predict the effect of a dose change, set `dose.amount` to the new dose while keeping `dose.input` as the original dose. PKNCA scales the concentrations proportionally.

```{r}
ss_half_dose <- superposition(
  o_conc_1,
  tau         = 24,
  dose.input  = actual_dose,
  dose.amount = actual_dose / 2,   # half dose
  n.tau       = Inf,
  check.blq   = FALSE
)

bind_rows(
  ss_converged   |> mutate(regimen = "Full dose"),
  ss_half_dose   |> mutate(regimen = "Half dose")
) |>
  ggplot(aes(x = time, y = conc, colour = regimen)) +
  geom_line() + geom_point(size = 2) +
  labs(title = "Dose scaling via superposition",
       x = "Time (h)", y = "Predicted SS concentration (mg/L)") +
  scale_colour_manual(values = c("Full dose" = "steelblue", "Half dose" = "coral")) +
  theme_minimal()
```

---

## Changing the dosing interval (τ)

Shortening τ to 8 h raises troughs and reduces peak-to-trough fluctuation:

```{r}
ss_8h <- superposition(
  o_conc_1,
  tau         = 8,
  dose.input  = actual_dose,
  dose.amount = actual_dose,
  n.tau       = Inf,
  check.blq   = FALSE
)

ggplot(ss_8h, aes(x = time, y = conc)) +
  geom_line(colour = "firebrick") +
  geom_point(size = 2, colour = "firebrick") +
  labs(title = "Superposition with τ = 8 h (more frequent dosing)",
       x = "Time within interval (h)", y = "Predicted SS concentration (mg/L)") +
  theme_minimal()
```

---

## Multiple doses within an interval

`dose.times` (default `0`) sets the dose time(s) within each interval, and a
vector places several doses inside one τ: `tau = 24` with
`dose.times = c(0, 12)` is twice-daily dosing described as a single 24 h
interval. `dose.amount` applies to **each** administration, so halving it here
keeps the total daily dose equal to the once-daily regimen above:

```{r}
ss_bid <- superposition(
  o_conc_1,
  tau         = 24,
  dose.input  = actual_dose,
  dose.amount = actual_dose / 2,  # half dose per administration
  dose.times  = c(0, 12),         # doses at 0 h and 12 h in each interval
  n.tau       = Inf,
  check.blq   = FALSE
)

bind_rows(
  ss_converged |> mutate(regimen = "Once daily, full dose"),
  ss_bid       |> mutate(regimen = "Twice daily, half dose")
) |>
  ggplot(aes(x = time, y = conc, colour = regimen)) +
  geom_line() + geom_point(size = 2) +
  labs(title = "Two doses per interval: dose.times = c(0, 12), τ = 24 h",
       x = "Time within interval (h)", y = "Predicted SS concentration (mg/L)",
       colour = NULL) +
  theme_minimal()
```

At the same total daily dose, splitting the dose lowers Cmax and raises the
trough. The dose times need not be evenly spaced: `dose.times = c(0, 8)` with
`tau = 24` describes a staggered morning/evening regimen (doses 8 h apart,
then a 16 h overnight gap).

---

## Extracting accumulation ratio

One common definition of the **accumulation ratio (Rac)** is Cmax at steady state divided by Cmax after the first dose (an AUC-based Rac, AUCτ,ss / AUCτ,first, is equally common):

```{r}
cmax_sd <- max(subj1$conc, na.rm = TRUE)
cmax_ss <- max(ss_converged$conc, na.rm = TRUE)
rac      <- cmax_ss / cmax_sd
cat("Accumulation ratio (Rac):", round(rac, 2), "\n")
cat("Single-dose Cmax:", round(cmax_sd, 2), "mg/L\n")
cat("Steady-state Cmax:", round(cmax_ss, 2), "mg/L\n")
```

---

## Key arguments reference

| Argument | Required | Meaning |
|---|---|---|
| `tau` | Yes | Dosing interval (same time units as data) |
| `dose.input` | No | Dose that generated the observed profile; supply together with `dose.amount` to dose-scale |
| `dose.amount` | No | Dose to simulate; if omitted, the output dose equals the input dose |
| `dose.times` | No (default `0`) | Dose time(s) within the interval; a vector gives multiple/staggered doses per τ |
| `n.tau` | No (default `Inf`) | Number of intervals to simulate before stopping |
| `steady.state.tol` | No (default 0.001) | Convergence tolerance for `n.tau = Inf` |
| `check.blq` | No (default `TRUE`) | Require first concentration = 0; set `FALSE` if first sample is non-zero |
| `auc.type` | No (default `"AUCinf"`) | How to extrapolate the single-dose tail: `"AUCinf"`, `"AUClast"`, or `"AUCall"` |
| `additional.times` | No | Extra timepoints to include in the output |

> **Limitation:** Superposition assumes linear PK. It should not be used for drugs with non-linear kinetics (e.g. Michaelis-Menten elimination, time-varying clearance, or saturable protein binding).


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [superposition()](https://humanpred.github.io/pknca/reference/superposition.html)
:::
