---
title: "Urine Excretion"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## Urine NCA overview

Urine NCA characterizes how much drug is excreted in urine over time. Key parameters:

| Parameter | Formula | Meaning |
|---|---|---|
| `ae` | Σ(concentration × volume) | Amount excreted in urine (mass) |
| `fe` | ae / dose | Fraction of dose excreted unchanged |
| `clr.obs` | ae / AUCinf.obs | Renal clearance — using observed Clast |
| `clr.pred` | ae / AUCinf.pred | Renal clearance — using predicted Clast |
| `clr.last` | ae / AUClast | Renal clearance — no extrapolation |

---

## Data structure for urine

Urine NCA requires a **separate concentration object** for urine, with the `volume` argument to `PKNCAconc()` specifying the column name that holds the urine volume collected in each interval.

The key distinction from plasma NCA: urine concentrations represent **interval averages** rather than timepoint measurements. Each row covers the collection period from the previous sample to the current time.

```{r}
# Simulate urine collection for 4 subjects after a 100 mg IV bolus
# Collections at: 0-2h, 2-6h, 6-12h, 12-24h
set.seed(99)

urine_collections <- expand.grid(
  Subject     = factor(1:4),
  end_time    = c(2, 6, 12, 24)
) |>
  mutate(
    start_time = c(0, 2, 6, 12)[match(end_time, c(2, 6, 12, 24))],
    # Simulated urine volume (mL) and concentration (mcg/mL)
    volume_mL  = round(runif(n(), 80, 300)),
    # Fraction excreted: ~30% in first 2h, 25% 2-6h, 20% 6-12h, 15% 12-24h
    ae_target  = 100 * c(0.30, 0.25, 0.20, 0.15)[match(end_time, c(2, 6, 12, 24))],
    ae_indiv   = ae_target * exp(rnorm(n(), 0, 0.1)),  # mild IIV
    conc_mcg_mL = ae_indiv / volume_mL * 1000           # conc = amount / volume
  ) |>
  select(Subject, time = end_time, conc = conc_mcg_mL, volume = volume_mL)

head(urine_collections)
```

```{r}
# Plasma concentrations (IV bolus, same subjects)
ke <- 0.1  # h^-1
d_plasma <- expand.grid(
  Subject = factor(1:4),
  time    = c(0.5, 1, 2, 4, 6, 8, 12, 16, 24)
) |>
  mutate(
    ke_i = ke * exp(rnorm(n(), 0, 0.15)),
    conc = 100 / 30 * exp(-ke_i * time)   # dose/V * e^(-ke*t)
  )

d_dose <- data.frame(
  Subject = factor(1:4),
  dose    = 100,    # mg
  time    = 0,
  route   = "intravascular"
)
```

---

## Setting up urine PKNCAconc

Pass the urine volume column name via the `volume` argument; the unit arguments (`concu`, `amountu`, `timeu`, …) are optional and set units. The formula uses the same `conc ~ time | Subject` structure.

```{r}
o_conc_plasma <- PKNCAconc(d_plasma, conc ~ time | Subject)
o_conc_urine  <- PKNCAconc(
  urine_collections,
  conc ~ time | Subject,
  volume = "volume"    # column with urine volume (mL)
)

o_dose <- PKNCAdose(d_dose, dose ~ time | Subject, route = "intravascular")
```

---

## Computing ae, fe, and renal clearance

Renal clearance combines the urine amount excreted (Ae) with the plasma AUC. Combined plasma+urine analysis in a single `PKNCAdata` object is limited; the recommended workflow is to run plasma NCA and urine NCA separately and join the results.

---

## Separate plasma + urine workflow

This separate-workflow approach works reliably:

```{r}
# --- Plasma NCA ---
o_plasma_data <- PKNCAdata(
  o_conc_plasma, o_dose,
  intervals = data.frame(start = 0, end = Inf,
                         auclast = TRUE, aucinf.obs = TRUE),
  impute = "start_conc0"
)
o_nca_plasma <- pk.nca(o_plasma_data)

plasma_auc <- as.data.frame(o_nca_plasma) |>
  filter(PPTESTCD %in% c("auclast", "aucinf.obs")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# --- Urine: compute ae manually ---
# ae = concentration (mcg/mL) × volume (mL) / 1000  →  mg
ae_by_subject <- urine_collections |>
  mutate(ae_interval = conc * volume / 1000) |>  # mcg/mL * mL / 1000 = mg
  group_by(Subject) |>
  summarise(ae_total = sum(ae_interval), .groups = "drop")

# --- Join and compute clr, fe ---
dose_amt <- 100  # mg

results <- plasma_auc |>
  left_join(ae_by_subject, by = "Subject") |>
  mutate(
    fe       = ae_total / dose_amt,
    clr_obs  = ae_total / aucinf.obs,
    clr_last = ae_total / auclast
  )

results
```

For a real IV analysis, back-extrapolate C0 instead — see the [Intravascular](intravascular.qmd) chapter; `conc = 0` at `t = 0` is used here only to keep the simulation short.

---

## Interpreting results

```{r}
results |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(Subject, ae_total, fe, clr_obs, clr_last) |>
  knitr::kable(
    col.names = c("Subject", "Ae (mg)", "fe", "CLr.obs (L/h)", "CLr.last (L/h)"),
    caption = "Urine NCA results — simulated IV bolus (dose = 100 mg)"
  )
```

**Interpretation guide:**

- `fe` near 0.30 means ~30% of dose is renally excreted unchanged. The remainder is metabolized or excreted by other routes.
- `clr.obs` ≈ GFR (120 mL/min = 7.2 L/h) suggests renal filtration; higher values indicate active secretion.
- `clr.last` > `clr.obs` because the same Ae is divided by the smaller AUClast; the gap grows when sampling misses late elimination.

---

## Using the PKNCA `ae` parameter natively

If your data are structured with urine as a separate concentration object and `PKNCAconc(..., volume = "vol_col")`, PKNCA can compute `ae` directly:

```{r}
# Structure for direct ae computation
o_conc_ur2 <- PKNCAconc(
  urine_collections,
  conc ~ time | Subject,
  volume = "volume"
)

ur_intervals <- data.frame(
  start = 0,
  end   = 24,
  ae    = TRUE
)

o_data_ae <- PKNCAdata(o_conc_ur2, o_dose, intervals = ur_intervals)
o_nca_ae  <- pk.nca(o_data_ae)

as.data.frame(o_nca_ae) |>
  filter(PPTESTCD == "ae") |>
  select(Subject, PPORRES) |>
  arrange(Subject)
```

---

## New urine parameters in PKNCA ≥ 0.12.2

### volpk — total collection volume

`volpk` is the sum of urine volumes across all collection periods within the analysis interval. When you pass `volume = "volume"` to `PKNCAconc()`, PKNCA can compute this automatically:

```{r}
vol_interval <- data.frame(
  start  = 0,
  end    = 24,
  ae     = TRUE,
  volpk  = TRUE    # sum of urine volumes over the interval
)

o_nca_vol <- pk.nca(PKNCAdata(o_conc_ur2, o_dose, intervals = vol_interval))
as.data.frame(o_nca_vol) |>
  filter(PPTESTCD %in% c("ae", "volpk")) |>
  select(Subject, PPTESTCD, PPORRES)
```

### Excretion rate parameters

| Parameter | Description |
|---|---|
| `ermax` | Maximum excretion rate (amount/time) within the interval |
| `ertmax` | Midpoint collection time of the maximum excretion rate period |
| `ertlst` | Midpoint collection time of the last measurable excretion rate |

These mirror the plasma-side `cmax` / `tmax` / `tlast` but apply to the excretion rate (amount excreted per collection period divided by collection duration).

```{r}
# Excretion rate needs each collection's duration, and PKNCA takes urine `time` as
# the collection START with `duration` extending forward — rebuild the object that way
urine_er <- urine_collections |>
  mutate(
    start_t  = c(0, 2, 6, 12)[match(time, c(2, 6, 12, 24))],
    duration = time - start_t
  ) |>
  transmute(Subject, time = start_t, conc, volume, duration)

o_conc_er <- PKNCAconc(urine_er, conc ~ time | Subject,
                       volume = "volume", duration = "duration")

er_interval <- data.frame(
  start  = 0,
  end    = 24,
  ae     = TRUE,
  ermax  = TRUE,    # maximum excretion rate
  ertmax = TRUE,    # midpoint time of the maximum-rate collection
  ertlst = TRUE     # midpoint time of the last measurable-rate collection
)

o_nca_er <- pk.nca(PKNCAdata(o_conc_er, o_dose, intervals = er_interval))
as.data.frame(o_nca_er) |>
  filter(PPTESTCD %in% c("ae", "ermax", "ertmax", "ertlst")) |>
  select(Subject, PPTESTCD, PPORRES)
```

### Dose-normalized renal clearance (≥ 0.12.2)

Dose-normalized renal clearance variants are now available in the parameter registry:

| Parameter | Normalized from |
|---|---|
| `clr.last.dn` | `clr.last / dose` |
| `clr.obs.dn` | `clr.obs / dose` |
| `clr.pred.dn` | `clr.pred / dose` |

Renal clearance divides the urine Ae by the plasma AUC — two different concentration objects — so a single `pk.nca()` call cannot compute `clr.*` directly. Compute each piece with `pk.nca()` and combine them, with `pk.calc.dn()` applying the dose normalization:

```{r}
o_nca_ae_dn <- pk.nca(PKNCAdata(o_conc_ur2, o_dose,
                                intervals = data.frame(start = 0, end = 24, ae = TRUE)))

clr_dn <- plasma_auc |>   # plasma pk.nca() results from the separate workflow above
  left_join(
    as.data.frame(o_nca_ae_dn) |>
      filter(PPTESTCD == "ae") |>
      select(Subject, ae = PPORRES),
    by = "Subject"
  ) |>
  mutate(
    ae          = ae / 1000,                  # native ae is in mcg (mcg/mL * mL); convert to mg to match the plasma AUC
    clr.last    = ae / auclast,
    clr.obs     = ae / aucinf.obs,
    clr.last.dn = pk.calc.dn(clr.last, 100),  # dose = 100 mg
    clr.obs.dn  = pk.calc.dn(clr.obs, 100)
  )

clr_dn
```


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html)
:::
